---
title: "Heart Disease Prediction"
author: "A. Nguyen, B. Waite"
date: "7/17/2019"
output:
  html_document: default
  pdf_document: default
---


#Introduction: 

This study hopes to identify factors contributing to the development of heart disease in an observed population and to train a logistic regression model to predict whether future patients will develop heart disease given the same set of factors. 

#Data: 

The data used in this study were obtained from the archives of the University of California, Irvine and are available on the web at: https://archive.ics.uci.edu/ml/datasets/Heart+Disease. 

A brief summary of the variables contained in the data is available in the table below. A more robust description of each variable is contained in Appendix A. 

```{r include = FALSE, message = FALSE}
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(GGally)

#Requires setwd() to the current directory with heart.csv
heart <- read_csv("heart.csv")

#Convert the categorical variables to factors
heart$sex = as.factor(heart$sex)
heart$cp = as.factor(heart$cp)
heart$fbs = as.factor(heart$fbs)
heart$restecg = as.factor(heart$restecg)
heart$exang = as.factor(heart$exang)
heart$slope = as.factor(heart$slope)
heart$ca = as.factor(heart$ca)
heart$thal = as.factor(heart$thal)
heart$target = as.factor(heart$target)

#Summary Statistcs
sumstats = summary(heart)

#Dataset Structure
datastructure = str(heart)
attach(heart)
```


```{r echo = FALSE}

vartable = as.data.frame(names(heart))
vartable$description = c("Age","Sex","Chest Pain Type","Resting Blood Pressure at Admission","Serum Cholesterol","Fasting Blood Sugar > 120 mg/dl","Resting Electrocardiographic Results","Maximum Heart Rate Achieved","Exercised Induced Angina", 
                         "ST Depression Induced by exercise relative to rest", "The Slope of the Peak Exercise ST Segment","Number of Major Vessels Colored by Flouroscopy", "Severity of Defect","Diagnosis of Heart Disease")
vartable$typeOrUnits = c("Years", "1 = Male;0 = Female", "0 = Typical Angina, 1 = Atypical Angina, 2 = Non-Anginal Pain, 3 = Asymptomatic", "mmHg","mg/dl","1 = true; 0 = false","0 = Normal, 1 = Having ST-T Wave Abnormality, 2 = Showing Probable or Definite left ventricular hypertrophy","Beats Per Minute (BPM)","1 = Yes; 0 = No","Unknown","1 = Upsloping, 2 = Flat, 3 = Downsloping","0-3","3 = normal; 6 = fixed defect; 7 = reversable defect","0 = No Heart Disease Diagnosed; 1 = Heart Disease Diagnosed")
vartable$usage = "Predictor"
vartable$usage[14] = "Response"
names(vartable) = c("Variable","Description","Units or Value","Usage")
kable(vartable) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


Some interesting breakdowns / observations contained within the data:

* The age range in the data set is from 29 to 77 years old. Age distribution is normal. 
* There are 303 total participants.  207 Male and 96 Female. 
* A general contingency table of Sex by Target (diagnosed with cardiac disease) shows: 

```{r echo = FALSE}
table(sex,target)
```

* In examining the histogram for Cholestorol values, we can see that there is at least one outlier with a value near 600 mg/dl. We might want to explore this further. 
* The oldpeak variable also shows some diferentiation in the outcome of having heart disease. 

```{r echo = FALSE}

#Boxplot of oldpeak
boxplot(oldpeak~target, main = "Heart Disease or Not vs Oldpeak", xlab = "Heart Disease Present", ylab = "Oldpeak Value")
```

* There also appear to be some outliers in the oldpeak data set that you can see on the boxplot above and are present in the histogram that is right skewed. Log transforming the oldpeak variable does help with normality. We don't know if this is required at this point, but is good to keep in mind and may help to improve our prediction model later. 

```{r echo = FALSE}
#Histogram of oldpeak
hist(oldpeak)
#Some strong right skew here that may need to be addressed if we use this as a continuous variable. And I think we should initially. 

#log transform of oldpeak
hist(log(oldpeak))
```

The maximum heart rate achieved during the stress test (thalach) also appears to be strongly correlated with the outcome of whether a patient will have heart disease or not. 

```{r echo = FALSE, message = FALSE} 
#Histogram of thalach
hist(thalach)

#Boxplot of thalach against response
boxplot(thalach~target, main = "Maximum Heart Rate Achieved Grouped by Heart Disease or No", xlab = "Heart Disease Present", ylab = "Heart Rate in Beats Per Minute (BPM)")
#Some good differentiation here
```

## Let's try to do some PCA: 



```{r include = FALSE}
#Get only the continuous variables
heartcont = heart[,c('age','trestbps','chol','thalach','oldpeak')]
pairs(heartcont, panel = panel.smooth)
```


Looking at these plots, there does not seem to be much correlation between the continuous variables. The heatmap below seems to confirm this. 

```{r echo = FALSE, message = FALSE}
library(gplots)
my.cor2 = cor(heartcont)
heatmap.2(my.cor2,col=redgreen(75), 
          density.info="none", trace="none", dendrogram=c("both"), 
          symm=F,symkey=T,symbreaks=T, scale="none",key=T)

cov(heartcont)
```

. 

```{r echo = FALSE}
#Looking at Principal Components with scaling (using the correlation values vice the covariance values)
pcomps = prcomp(heartcont,scale. = TRUE)
pcscores = pcomps$x
pairs(pcscores, panel = panel.smooth)
```

There do not seem to be any correlations in the Principal Components. Let's try to produce the Scree / Cumulative Proportion Plots: 

```{r echo = FALSE}
par(mfrow=c(1,2))
eigenvals<-(pcomps$sdev)^2
plot(1:5,eigenvals/sum(eigenvals),type="l",main="Scree Plot",ylab="Prop. Var. Explained", ylim = c(0,.4))
cumulative.prop<-cumsum(eigenvals/sum(eigenvals))
plot(1:5,cumulative.prop,type="l",main="Cumulative proportion")# ,ylim=c(0,1))
par(mfrow=c(1,1))
```

It looks like from the above plot, that approximately 90% of the variability can be explained using 4 principal components. It does appear, though that all 5 continue to contribute to the model which is consistent with our earlier assessment that there is not significant correlation among the predictors. 

```{r echo = FALSE, message = FALSE}
pcscores = data.frame(pcscores)
ggplot(data = pcscores, aes(x = PC1, y = PC2)) +
  geom_point(aes(col=target), size=1)+
  ggtitle("PCA of Heart Disease Data")

heartcont$target = as.factor(heart$target)
```

This pairs plot explores the correlation between the continuous predictors grouped by the outcome: 

```{r echo = FALSE, message = FALSE}
#pairs on variables grouped by target

ggpairs(heartcont, columns=1:5, aes(color=target)) + 
  ggtitle("Heart Disease Original Variables Grouped by Target (Diagnosis)")
```

While this pairs plot explores the correlation between the Principal Components grouped by the outcome. 

```{r echo = FALSE, message = FALSE}
#pairs on principal components grouped by target
pcscores$target = as.factor(heart$target)
ggpairs(pcscores, columns=1:5, aes(color=target)) + 
  ggtitle("Heart Disease Principal Components Grouped by Target (Diagnosis)")

```

Running a number of PCA charts coded by response does not appear to show a clean discriminant that we can use to do regression on the Principal Components. Principal component 1 (PC1) does show some ability to distinguish against PC2, but there is still significant overlap. This could mean that our logistic model will not perform super well. It could also mean that we just don't have very many continuous predictors and that the categorical predictors will really help us differentiate with interaction terms. 

Looking at the weighting on the principal components: 

```{r echo = FALSE}
pcomps$rotation
```

In looking at this table, we can see that each variable contributes a significant amount in at least one of the Principal Components. This is a good indication that the variables are not correlated with each other and that each might contribute something unique to our model. 

# Prep for Analysis

I considered doing a Train / Test split on the data, but think that we might try to do a k-fold cross validation of the model. In order to do that, we have to train the model first. 

```{r echo = FALSE, message = FALSE}
library(caret) 

#define the training control method for the model
train_control = trainControl(method = "cv", number = 10)

#fit the simple model using all the variables
simple_model = train(target~., data = heart, trControl = train_control, method = "glm", family = "binomial")
print(simple_model)
```

and the coefficients: 

```{r echo = FALSE, message = FALSE}
simple_model$finalModel
```

Note that we didn't do any feature selection here, so let's try to add that in with Stepwise AIC based feature selection: 

```{r echo = FALSE, message = FALSE}
#This line makes a lot of output when knitted
simple_model_stepwise = train(target~., data = heart, trControl = train_control, method = "glmStepAIC", family = "binomial")

simple_model_stepwise$finalModel$coefficients

simple_model_stepwise$results
```

Just to try another selection algorithm, used "LogitBoost" from the Caret library. It doesn't consistently outperform the simpler stepwise selection or "all-in" methods. We haven't learned the interpretation for this model.  

```{r echo = FALSE, message = FALSE}
simple_model_boost = train(target~., data = heart, trControl = train_control, method = "LogitBoost", family = "binomial")

simple_model_boost$finalModel

simple_model_boost$results
```


# Appendix A: 


Variable names and meanings pulled from the source website: https://archive.ics.uci.edu/ml/datasets/Heart+Disease


*Chest Pain Type (cp)* [SOURCE: https://www.aafp.org/afp/2005/1115/p2012.html]
  
Angina is chest pain that results from reduced blood flow to the heart muscle.

Typical angina (definite chest pain) meets three of the following characteristics

1. Substernal chest discomfort of characteristic quality and duration
2. Provoked by exertion or emotional stress
3. Relieved by rest and/or nitroglycerine

Atypical angina (probable chest pain) meets two of the characteristics

Non-anginal pain meets one characteristic

Asymptomatic means no chest pain
    

*Resting Blood Pressure (trestbps)*

High blood pressure is a common condition and major risk factor for heart disease. Uncontrolled hypertension can lead to heart attack and stroke
    
*Serum Cholesterol (chol)* 

[SOURCE: https://www.medicalnewstoday.com/articles/321519.php]

Serum cholesterol measures the amount of triglycerides, low-density lipoprotein (LDL/bad cholesterol), and high density lipoprotein (HDL/good cholesterol). High serum cholesterol levels are associated with higher risk for coronary artery disease and heart attack

*Normal Ranges for cholesterol levels*

<= 19 years old - serum chol [<= 170 mg/dL] - HDL [>= 45 mg/dL] - LDL [< 100 mg/dL] - Triglycerides [< 150 mg/dL]
>= 20 years old female - serum chol [125-200 mg/dL] - HDL [>= 50 mg/dL] - LDL [< 100 mg/dL] - Triglycerides [< 150 mg/dL]
>= 20 years old male - serum chol [125-200 mg/dL] - HDL [>= 40 mg/dL] - LDL [< 100 mg/dL] - Triglycerides [< 150 mg/dL]

*Fasing Blood Sugar (fbs)* 

[SOURCE: https://www.mayoclinic.org/diseases-conditions/diabetes/diagnosis-treatment/drc-20371451]

Ranges:

1. fbs < 100 mg/dL is normal
2. fbs 100-125 mg/dL is prediabetes
3. fbs >= 126 mg/dL on two separate tests diagnoses diabetes

A fasting blood sugar level of more than 125 mg/dL is current threshold for diabetes. Higher fasting blood sugars are associatied with higher risk for heart disease. 
    
*Resting Electrocardiographic Results (restecg)* 

[SOURCE: https://www.mayoclinic.org/diseases-conditions/left-ventricular-hypertrophy/symptoms-causes/syc-20374314]

ST-T Wave Abnormality (1) are associated with increased cardiac deaths and are highly suggestive of coronary artery disease when associcated with cardiac signs and symptoms. T wave inversions and/or ST elevation or depression of > 0.05 mV. 

Left Ventricular Hypertrophy (2) is enlargement/thickening (hypertrophy) of the walls of your heart's main pumping chamber (left ventricle) and causes it to work harder. 

  * Defined using Estes' criteria
  * As the workload increases, the muscle tissue in the chamber wall thickens. The enlarged heart muscle loses elasticity and may eventually fail to pump with as much force as needed.
  * Left ventricular hypertrophy is more common in people who have uncontrolled high blood pressure and puts you at higher risk of a heart attack and stroke.
      
  
*Maximum Heart Rate Achieved (thalach)* 

[SOURCE: https://www.hopkinsmedicine.org/health/wellness-and-prevention/understanding-your-target-heart-rate]

The maximum heart rate is based on your age, as subtracted from 220. 

  * For a 50-year-old, maximum heart rate is 220 minus 50, or 170 beats per minute. 
  * Target heart rate that a 50-year-old would want to aim for during exercise is 85 to 145 beats per minute (50%-85% exertion level).
  * Maximum heart rates below these averages may indicate weakening of heart muscle
    
*Exercise Induced Angina*

Chest pain that is triggered by physical exertion. When you climb stairs, exercise or walk, your heart demands more blood, but it's harder for the muscle to get enough blood when your arteries are narrowed.
    
*ST Depression Induced (oldpeak)*

[SOURCE: https://www.ahajournals.org/doi/pdf/10.1161/01.CIR.80.1.87]

* Units are in mm or mV?

 ST segment depression in PQRST wave (ECG) may be assoicated with coronary artery disease with larger depressions indicative of abnormality. Exercise stress test requires an ST depression of at least 2 mm to significantly indicate reversible ischaemia [WIKI]
  
*Slope of Peak Exercise ST Segment (slope)*

[SOURCE: https://ecgwaves.com/ecg-topic/exercise-stress-test-ecg-symptoms-blood-pressure-heart-rate-performance/]

* Steep slopes indicate a less likely chance of ischemia
* Horizontal/Flat slopes indicate a more likely chance of ischemia
  
*Number of Major Vessels Colored by Flouroscopy (ca)*

*Severity of Defect (thal)*

*target*

Diagnosis of heart disease (angiographic disease status)

Value 0: < 50% diameter narrowing
Value 1: > 50% diameter narrowing (in any major vessel: attributes 59 through 68 are vessels)

# Appendix B: Additional Exploratory Analysis Performed

```{r include = FALSE}
#Age distribution
hist(age)

#Heart Disease or not by Age
boxplot(age~target, main = "Heart Disease or Not by Age", xlab = "Heart Disease Present", ylab = "Age in Years")


#Sex breakdown
kable(table(sex))


#Chest pain breakdown
kable(table(cp))
```

 

```{r}
#Histogram of trestbps
hist(trestbps)

#Boxplot of trestbps by response
boxplot(trestbps~target, main = "Heart Disease or Not by trestbps", ylab = "Beats Per Second", xlab = "Heart Disease Present")

#Histogram of chol
hist(chol)

#Boxplot of chol vs response
boxplot(chol~target, main = "Heart Disease or Not by chol", xlab = "Heart Disease Present", ylab = "Cholestorol in mg/dl")

#Values of fbs in the data set
kable(table(fbs))

#Values of restecg in the data set
kable(table(restecg))

#Histogram of thalach
hist(thalach)

#Boxplot of thalach against response
boxplot(thalach~target)
#Some good differentiation here

#Values of exang in the data
kable(table(exang))

#Histogram of oldpeak
hist(oldpeak)
#Some strong right skew here that may need to be addressed if we use this as a continuous variable. And I think we should initially. 

#Boxplot of oldpeak
boxplot(oldpeak~target, main = "Heart Disease or Not vs oldpeak")

#Values of slope occuring
kable(table(slope))

#Values of ca occuring
kable(table(ca))

#Values of thal occuring
kable(table(thal))

#Distribution of the target response seems roughly 50-50
kable(table(target))

```